From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-24 Pi/paper/examples.tex
fixed example in Sec. 2

For concreteness, we demonstrate two different circuits that implement the
following reversible function specification
$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2),
~b_1, ~b_2)$ where $\vee$ is boolean disjunction and $\underline{\vee}$ is the
exclusive-or operation. The circuits are presented in both the textual interface
\verb|qasm| and the graphical interface:

@@

There is a wealth of manual and algorithmic approaches for producing circuits such
as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left
was manually produced using a standard synthesis algorithm for reversible
circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced
using an approach that analyzes the recursive structure of the circuit (and
would generalise to computing the disjunction of more than two inputs):

@@

Although rather trivial, the examples above illustrate the general idea. A more
interesting example would be the classical core of Shor's algorithm which
requires a circuit implementing $f(r) = a^{r} \mod N$ for fixed $a$ and $N$. The
specification of the circuit is relatively straightforward to calculate. Here it
is for $a=11$ and $N=15$:
\[\begin{array}{rcll}
g(r,h) &=& \left\{ \begin{array}{ll}
                     (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
                     (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
                     (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
                     (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
                                \end{array}\right.
\end{array}\]

\noindent However, as explained in standard accounts of the algorithm (e.g., the
Qiskit implementation), producing an efficient modular exponentiation circuit
from this specification is not straightforward and is actually the bottleneck in
Shor’s algorithm. Typical derivations of the circuit start from elementary
gates, build a circuit for reversible disjunction (like the two circuits above),
reversible conjunction, a circuit for a half-adder, a circuit for computing the
carry, progressing to a circuit for modular addition, which is used to build a
circuit for modular multiplication, and then finally a circuit for modular
exponentiation taking care at each step to avoid the exponential blowup (e.g.,
by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.

>>>>

For concreteness, we demonstrate two different circuits that implement the
following reversible function specification
$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2),
~b_1, ~b_2)$ where~$\vee$ is disjunction and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface
\verb|qasm| and the graphical interface:

@@

There is a wealth of manual and algorithmic approaches for producing circuits
such as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the
left was manually produced using a standard synthesis algorithm for reversible
circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced
using an approach that analyzes the recursive structure of
$\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
more than two inputs):

@@

Although rather trivial, the examples above illustrate the general idea. A more
interesting example would be the classical core of Shor's algorithm which
requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$ for
fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a
reversible specification of this function is $f^R(r,h) = (r, f(r)
~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is
applied bitwise. When $h=0$ the second component of the result of $f^R$ is the
result of the original modular exponentiation $f$.
%%
%% the specification of the circuit is relatively straightforward to
%% calculate. Here it is for $a=11$ and $N=15$:
%% \[\begin{array}{rcll}
%% g(r,h) &=& \left\{ \begin{array}{ll}
%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
%%                                 \end{array}\right.
%% \end{array}\]
%%
However, as explained in standard accounts of the algorithm (e.g., the
Qiskit implementation), producing an efficient modular exponentiation circuit
from this specification is not straightforward and is actually the bottleneck in
Shor’s algorithm. Typical derivations of the circuit start from elementary
gates, build a circuit for reversible disjunction (like the two circuits above),
reversible conjunction, a circuit for a half-adder, a circuit for computing the
carry, progressing to a circuit for modular addition, which is used to build a
circuit for modular multiplication, and then finally a circuit for modular
exponentiation taking care at each step to avoid the exponential blowup (e.g.,
by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.
